# National Storm and Severe Weather Data Analysis

**NOTE**: *This analysis is created as a class assignment for Coursera's "Reproducible Research" class, from Data Science track, offered by Dr. Roger Peng.*

## Summary
This report evaluates the effects of different types of natural events based on their impacts on human health and economy. The data used for this analysis comes from US National Oceanic and Atmospheric Administration's (NOAA) storm database. After considering data from 1950 to 2011, we conclude that tornados caused the most fatalities and injuries over the considered time period accross US. In terms of economic impact, floods have been the costliest causing the most property and crop damage.

## Data
The data for this assignment is in CSV format compressed using bzip2 algorithm. The code downloads it from National Weather Service's website ([link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)). The documentation for data is available [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) and [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf).

## Data Processing
### Step 1: Download Data
```{r message=FALSE}
#Setup
library(R.utils)
library(reshape2)
library(ggplot2)

#Download Data
filePath <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if (!file.exists("StormData.csv.bz2")){
        download.file(filePath, destfile = "StormData.csv.bz2", method = "curl")} #method = "curl" because the site is https
if (!file.exists("StormData.csv")){
        bunzip2("StormData.csv.bz2", "StormData.csv", remove = F)}
```

### Step 2: Load Data and Create Aggregates
Here's what that raw data looks like
```{r cache=TRUE}
data <- read.csv("StormData.csv")
str(data)
```

We considered data range 1950 - 2011. Health impact is computed by summing `FATALITIES` and `INJURIES` by `EVTYPE` variable.
```{r}
health <- aggregate(cbind(FATALITIES, INJURIES) ~ EVTYPE, data = data, sum)
property <- aggregate(PROPDMG ~ EVTYPE + PROPDMGEXP, data = data, sum)
crop <- aggregate(CROPDMG ~ EVTYPE + CROPDMGEXP, data = data, sum)
```

### Step 3: Analyze Health Impact
5 deadliest event types and the injuries and fatalities they caused are as follows
```{r}
#Identify 5 most harmful events
health <- health[order(health$FATALITIES, health$INJURIES, decreasing = T),]
health$EVTYPE <- factor(as.character(health$EVTYPE), levels = as.character(health$EVTYPE))
TOTAL <- health$FATALITIES + health$INJURIES
health <- cbind(health, TOTAL)
health1 <- health[1:5, ]
id <- 1:5
health1 <- cbind(id, health1)

#Melt the dataframe with one variable value appened to other on variable
health1<-melt(health1[c('EVTYPE','FATALITIES','INJURIES')], id.vars=1)
```

### Step 4: Analyze Economic Impact
We now use crop damage and property damage currency amounts to evaluate economic impact. `PROPDMG` and `CROPDMG` are expressed in different measures of units (`PROPDMGEXP` and `CROPDMGEXP` respectively). Here are the different measure of units in for these two measurements in the data.
```{r}
table(property$PROPDMGEXP)
table(crop$CROPDMGEXP)
```

**Interpretation of Units:** k=K=10^3, m=M=10^6, h=H=10^2, i=10^1.  
**Proposed Common Unit of Currency:** k=K=10^3 (Dollar Amount in Thousand)  

Here's a function that converts all measures of unit to k=K=10^3 for all `PROPDMG` and `CROPDMG` records.
```{r}
unify <- function(economy, colValue, colUnit) {
    economy1 <- economy[!economy[[colUnit]] %in% c("", "-", "?", "+"), ]
    economy1[[colValue]] <- as.numeric(economy1[[colValue]])
    value.In.K <- numeric()
    for (i in 1:nrow(economy1)) {
        oriNum <- economy1[i, ][[colValue]]
        if (economy1[i, ][[colUnit]] == "0") {
            damage <- oriNum/1000
        } else if (economy1[i, ][[colUnit]] == "1") {
            damage <- oriNum/100
        } else if (economy1[i, ][[colUnit]] == "2") {
            damage <- oriNum/10
        } else if (economy1[i, ][[colUnit]] == "3") {
            damage <- oriNum
        } else if (economy1[i, ][[colUnit]] == "4") {
            damage <- oriNum * 10
        } else if (economy1[i, ][[colUnit]] == "5") {
            damage <- oriNum * 100
        } else if (economy1[i, ][[colUnit]] == "6") {
            damage <- oriNum * 1000
        } else if (economy1[i, ][[colUnit]] == "7") {
            damage <- oriNum * 10^4
        } else if (economy1[i, ][[colUnit]] == "8") {
            damage <- oriNum * 10^5
        } else if (economy1[i, ][[colUnit]] == "B") {
            damage <- oriNum * 10^6
        } else if (economy1[i, ][[colUnit]] == "H" | economy1[i, ][[colUnit]] == 
            "h") {
            damage <- oriNum/10
        } else if (economy1[i, ][[colUnit]] == "M" | economy1[i, ][[colUnit]] == 
            "m") {
            damage <- oriNum * 10^3
        } else {
            damage <- oriNum
        }
        value.In.K <- c(value.In.K, damage)
    }
    economy1 <- cbind(economy1, value.In.K)
    economy1 <- aggregate(value.In.K ~ EVTYPE, data = economy1, sum)
    economy1 <- economy1[order(economy1$value.In.K, decreasing = T), ] #Order by Econmic Impact Value
    invisible(economy1)
}
```

Convert measures of unit.
```{r}
property1 <- unify(property, "PROPDMG", "PROPDMGEXP")
crop1 <- unify(crop, "CROPDMG", "CROPDMGEXP")
```

Merge currency adjusted property and crop damage datasets into a `economy` dataset.
```{r}
merge <- function(property, crop) {
    economy <- rbind(property, crop)
    economy <- aggregate(value.In.K ~ EVTYPE, data = economy, sum)
    economy <- economy[order(economy$value.In.K, decreasing = T), ]
    names(economy) <- c("EVTYPE", "Total.Value")
    Property.Value <- numeric(length = nrow(economy))
    Crop.Value <- numeric(length = nrow(economy))
    economy <- cbind(economy, Property.Value, Crop.Value)
    for (i in 1:nrow(economy)) {
        type <- economy[i, ]$EVTYPE
        if (type %in% property$EVTYPE) {
            economy[i, ]$Property.Value <- property[property$EVTYPE == type, 
                ]$value.In.K
        }
        if (type %in% crop$EVTYPE) {
            economy[i, ]$Crop.Value <- crop[crop$EVTYPE == type, ]$value.In.K
        }
    }
    invisible(economy)
}
economy <- merge(property1, crop1)
```

Identify 5 economically damading events
```{r}
#Identify 5 economically damaging events
economy1 <- economy[1:5, ]
id <- 1:5
economy1 <- cbind(id, economy1)

#Melt the dataframe with one variable value appened to other on variable
economy1<-melt(economy1[c('EVTYPE','Property.Value','Crop.Value')], id.vars=1)
```

## Results
From the following plot, it is clear that tornados have been deadlist, causing most harm to humans.
```{r}
#Plot
p <- ggplot(health1, aes(EVTYPE, value, fill = variable, ymax=10000)) + geom_bar(aes(order = value), stat="identity", position="dodge")
p <- p+geom_text(aes(label = value), vjust=-0.25, position=position_dodge(0.9), size = 3)
p <- p+ggtitle("5 Deadliest Event Types in United States from 1950 to 2011")
p <- p+theme(plot.title = element_text(lineheight=.8, face="bold"), axis.text.x = element_text(angle = 90, hjust = 1))
p <- p+xlab("Event Type")+ylab("Number of Humans Affected")+labs(fill="Types of Human Harm")
print(p)
```

From the following plot, it is clear that floods have caused the most economic damage, both to property as well as crops.
```{r}
#Plot
q <- ggplot(economy1, aes(EVTYPE, value, fill = variable, ymax=1000000)) + geom_bar(aes(order = value), stat="identity", position="dodge")
q <- q+geom_text(aes(label = value), vjust=-0.25, position=position_dodge(0.9), size = 3)
q <- q+ggtitle("5 Costliest Events in United States from 1950 to 2011")
q <- q+theme(plot.title = element_text(lineheight=.8, face="bold"), axis.text.x = element_text(angle = 90, hjust = 1))
q <- q+xlab("Event Type")+ylab("USD Amount in Thousands")+labs(fill="Types of Economic Impact")
print(q)
```